---
title: "Data Exploration, Amelia Grant-Alfieri"
output: html_document
---

## Evaluate the relationship between PM 2.5 ambient air concentration in the 5 New York City boroughs and percentage rate of asthma emergency room visits and the percentage rate of heart attack hospital admissions. 

```{r message=FALSE, output=FALSE}
library(tidyverse)
library(purrr)
library(broom)
library(rvest)
library(httr)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
getwd()
```

### Create PM2.5 Dataset, 2014
```{r message=FALSE, output=FALSE}
PM = read_csv("PM2.5_2014.csv") %>%
  janitor::clean_names() %>%
  #filter(borough == "Bronx" | borough == "Brooklyn" | borough == "Queens" | borough == "Staten Island" | borough == "Manhattan") %>%
  #filter(geo_type_name == "Borough") %>%
  #select(-geo_type_name, -geography, -geography_id, -indicator_name) %>%
  rename(PM_mean = mean_mcg_per_cubic_meter, 
         PM_tenth_percentile = x10th_percentile_mcg_per_cubic_meter,
         PM_ninety_percentile = x90th_percentile_mcg_per_cubic_meter, 
         PM_year = year,
         county_name = borough) %>%
  mutate(county_name = recode(county_name, "Brooklyn" = "Kings", "Staten Island" = "Richmond", "Manhattan" = "New York"))
```

### Create Children's Asthma Hospitalization Rate Dataset, 2012-14
```{r message=FALSE, output=FALSE}
asthma_hosp_child = read_csv("asthma_hosp_rate_10000_age0-17.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years) %>%
  rename(asthma_hosp_percent_rate = percent_rate, 
         asthma_hosp_years = data_years)
```

### Create Age-Adjusted Cardiovascular Hospitalization Rates Dataset, 2012-14
```{r message=FALSE, output=FALSE}
aa_CVD_hosp = read_csv("age_adj_CVD_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(health_topic == "Cardiovascular Disease Indicators") %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years) %>%
  rename(aa_CVD_percent_rate = percent_rate, 
         aa_CVD_years = data_years)
```

### Create Age-Adjusted Heart Attack Hospitalization Rates Dataset, 2014
```{r message=FALSE, output=FALSE}
aa_heart_attack_hosp = read_csv("age_adj_heart_attack_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(aa_hattack_percent_rate = percentage_rate_ratio,
         aa_hattack_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

### Create Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

### Create PM2.5 Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
PM_county_NYS = read_csv("PM2.5_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

### Plot PM2.5 (ug/m3) over time, 2000-2014
```{r}
PM_plot_conc_allyears = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Ambient PM2.5 Concentrations in New York State, 2000-2014", x = "Year", y = "PM2.5 (ug/m3)")
PM_plot_conc_allyears
```
### Plot PM2.5 (person days) over time, 2000-2014
```{r}
PM_plot_days_allyears = PM_county_NYS %>%
  filter(measure == "Person Days") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Time at Risk for Ambient PM2.5 Exposure in New York State, 2000-2014", x = "Year", y = "PM2.5 (person days)")
PM_plot_days_allyears
```


### Create Ozone Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
ozone_county_NYS = read_csv("ozone_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

### Plot Ozone over time, 2000-2014
```{r}
ozone_plot_allyears = ozone_county_NYS %>%
  filter(measure == "Person Days") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Time at Risk of Ambient Ozone Exposure in New York State, 2000-2014", x = "Year", y = "Ozone (person days)")
ozone_plot_allyears
```
### Plot Ozone per county, 2014
```{r}
ozone_plot_14 = ozone_county_NYS %>%
  filter(measure == "Person Days") %>%
  filter(year == "2014") %>%
  group_by(county) %>%
  ggplot(aes(x = reorder(county, -output), y = output, group = 1)) + geom_histogram(stat = "identity") + labs(title = "Time at Risk of Ambient Ozone Exposure in New York State, 2014", x = "County", y = "Ozone (person days)") + theme(axis.text.x = element_text(angle = 90))
ozone_plot_14
```

### Plot PM2.5 per county, 2014
```{r}
PM_plot_14 = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  filter(year == "2014") %>%
  group_by(county) %>%
  ggplot(aes(x = reorder(county, -output), y = output, group = 1)) + geom_histogram(stat = "identity") + labs(title = "Ambient PM2.5 Concentrations in New York State, 2014", x = "County", y = "PM2.5 (ug/m3)") + theme(axis.text.x = element_text(angle = 90))
PM_plot_14
```

### Plot Heart Attack Hospital Visits per County, 2014
```{r}
heart_plot_14 = aa_heart_attack_hosp %>%
  ggplot(aes(x = reorder(county, -aa_hattack_percent_rate), y = aa_hattack_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Heart Attack Hospital Visit Rate per 10,000 by County in New York State, 2014")
heart_plot_14 
```

### Plot Asthma ER Admissions per County, 2014
```{r}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), y = asthma_ER_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Asthma ER Admission Rate (per 10,000)", title = "Asthma Emergency Room (ER) Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```






### Join heart attack, asthma, and PM datasets
redo join by county. not all counties overlap in all datasets!!!
```{r}
to_join_asthma_14 = asthma_ER %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex")

to_join_heart_14 = aa_heart_attack_hosp %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex")
  
to_join_ozone_14 = ozone_county_NYS %>%
  filter(year == "2014") %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex") %>%
  rename(ozone_reading = output, ozone_unit = measure)

to_join_PM_14 = PM_county_NYS %>%
  filter (year == "2014") %>%
  rename (PM_reading = output, PM_unit = measure)

PM_ozone_joined = inner_join(to_join_ozone_14, to_join_PM_14, by = "county")

heart_asthma_joined = inner_join(to_join_heart_14, to_join_asthma_14, by = "county")

joined = inner_join(heart_asthma_joined, PM_ozone_joined, by = "county") %>%
  select(county, aa_hattack_percent_rate, asthma_ER_percent_rate, ozone_reading, ozone_unit, PM_reading, PM_unit)
```

### Plot Heart Attack & PM2.5, 2014
```{r}
heart_PM_plot = joined %>%  
  filter(PM_unit == "Micrograms/cubic meter (LC)") %>%
  group_by(county) %>%
  ggplot(aes(x = PM_reading, y = aa_hattack_percent_rate, color = county)) + geom_point() + labs(x = "", y = "", title = "by County in New York State, 2014") + labs(x = "Ambient PM2.5 (ug/m3)", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Association between PM2.5 and Heart Attack Hospital Visits in New York State, 2014")
heart_PM_plot
```

## Plot Asthma & PM2.5, 2014
```{r}
asthma_PM_plot = joined %>% 
  filter(PM_unit == "Micrograms/cubic meter (LC)") %>%
  group_by(county) %>%
  ggplot(aes(x = PM_reading, y = asthma_ER_percent_rate, color = county)) + geom_point() + labs(x = "Ambient PM2.5 (ug/m3)", y = "Asthma ER Admission Rate (per 10,000)", title = "Association between PM2.5 and Asthma Emergency Room (ER) Admissions in New York State, 2014")
asthma_PM_plot
```

## Ploto Heart Attack & Ozone, 2014
```{r}
heart_ozone_plot = joined %>%  
  filter(ozone_unit == "Person Days") %>%
  group_by(county) %>%
  ggplot(aes(x = ozone_reading, y = aa_hattack_percent_rate, color = county)) + geom_point() + labs(x = "Ozone (person days)", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Association between Ozone and Heart Attack Hospital Visits in New York State, 2014")
heart_ozone_plot
```

## Plot Asthma & Ozone, 2014
```{r}
asthma_ozone_plot = joined %>% 
  filter(ozone_unit == "Person Days") %>%
  group_by(county) %>%
  ggplot(aes(x = ozone_reading, y = asthma_ER_percent_rate, color = county)) + geom_point() + labs(x = "Ozone (person days)", y = "Asthma ER Admission Rate (per 10,000)", title = "Association between Ozone and Asthma Emergency Room (ER) Admissions in New York State, 2014")
asthma_ozone_plot
```

