---
title: "Data Exploration, Amelia Grant-Alfieri"
output: html_document
---

## Evaluate the relationship between PM 2.5 ambient air concentration in the 5 New York City boroughs and percentage rate of asthma emergency room visits and the percentage rate of heart attack hospital admissions. 

```{r message=FALSE, output=FALSE}
library(tidyverse)
library(purrr)
library(broom)
library(rvest)
library(httr)
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
getwd()
```

### Create PM2.5 Dataset, 2014
```{r message=FALSE, output=FALSE}
PM = read_csv("PM2.5_2014.csv") %>%
  janitor::clean_names() %>%
  filter(borough == "Bronx" | borough == "Brooklyn" | borough == "Queens" | borough == "Staten Island" | borough == "Manhattan") %>%
  filter(geo_type_name == "Borough") %>%
  select(-geo_type_name, -geography, -geography_id, -indicator_name)
```

### Create Children's Asthma Hospitalization Rate Dataset, 2012-14
```{r message=FALSE, output=FALSE}
asthma_hosp_child = read_csv("asthma_hosp_rate_10000_age0-17.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years)
```

### Create Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years)
```

### Create Age-Adjusted Cardiovascular Hospitalization Rates Dataset, 2012-14
```{r message=FALSE, output=FALSE}
aa_CVD_hosp = read_csv("age_adj_CVD_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years)
```

### Create Age-Adjusted Heart Attack Hospitalization Rates Dataset, 2014
```{r message=FALSE, output=FALSE}
aa_heart_attack_hosp = read_csv("age_adj_heart_attack_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years)
```

### Plot Heart Attack Hospital Visits per Borough
```{r}
heart_attack = aa_heart_attack_hosp %>%
  ggplot(aes(x = county_name, y = percentage_rate_ratio, group = 1)) + geom_histogram(stat = "identity")
heart_attack
```

### Plot Asthma ER Admissions per Borough
```{r}
asthma = asthma_ER %>%
    ggplot(aes(x = county_name, y = percentage_rate_ratio, group = 1)) + geom_histogram(stat = "identity") 
asthma 
```

### Plot PM 2.5 levels per Borough
```{r}
PM = PM %>%
ggplot(aes(x = borough, y = mean_mcg_per_cubic_meter, group = 1)) + geom_histogram(stat = "identity")
PM
```

### Join heart attack, asthma, and PM datasets
```{r}

```

### Plot Heart Attack & PM2.5, 2014
```{r}
#heart_attack_PM =  %>%  #use TBD joined dataframe
#ggplot(aes(x = borough, y = mean_mcg_per_cubic_meter, group = 1)) + geom_histogram(stat = "identity")
#heart_attack_PM
```

## Plot Asthma & PM2.5, 2014
```{r}
#asthma_PM =  %>% #use TBD joined dataframe
#ggplot(aes(x = borough, y = mean_mcg_per_cubic_meter, group = 1)) + geom_histogram(stat = "identity")
#asthma_PM
```

# Questions:

What to do about age-adjusted heart attack admissions but not age-adjusted asthma admissions?
