---
title: "Explore"
author: "Divya Bisht"
date: "12/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Early exploration was done with data from New York City counties and NY state-wide counties 

Looking at all age groups instead of stratifying by age because trends of exposure were similar across age groups. 



There was limited data available for annual PM2.5 by NYC counies, so NY state-wide plots were selected to demonstrate the relationship between PM2.5 and hospitalizations.  


Over 2000-2014, levels of PM2.5 across all counties in New York State have steadily decreased. 

### PM2.5 Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
PM_county_NYS = read_csv("PM2.5_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

### Plot PM2.5 (ug/m3) over time, 2000-2014 
```{r}
PM_plot_conc_allyears = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Ambient PM2.5 Concentrations in New York State, 2000-2014", x = "Year", y = "PM2.5 (ug/m3)")
PM_plot_conc_allyears
```


Next, we wanted to assess how PM2.5 exposure varied across counties in NY State (USE ANJILE'S exploratory) 

```{r}

```


Literature shows that exposure to PM2.5 is associated with asthma and cardiovascular disease. Therefore, we wanted to assess hospitalization of asthma and CVD across counties. 

### Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

### Plot Asthma ER Admissions per County, 2014
```{r}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), y = asthma_ER_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Asthma ER Admission Rate (per 10,000)", title = "Asthma Emergency Room (ER) Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```

Looking at cardiovascular hospitalizations...
* Data source: NYSDOH Health Data
```{r load_clean_cvd, message = FALSE}
cvd_data = read_csv(file = "./data_AK/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate)
```

### Age-Adjusted CVD Hospitalization Rate in NY State, 2012 - 2014 

```{r cvd_bar_nys}
cvd_data %>%
  ggplot(aes(x = reorder(county_name, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")
```



