---
title: "Explore"
author: "Divya Bisht"
date: "12/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## To be copy/pasted into the Explore tab on the Ashley's website project. 

Early exploration was done with data from New York City counties and NY state-wide counties 

Looking at all age groups instead of stratifying by age because trends of exposure were similar across age groups

(Insert: 
- Anjile's asthma hospitalizations by NYC borough - FIND THE DATASET 
- Ashley's cariovascular FIND the dataset (ageadjusted_cvd_hospitalizations)
- Amelia PM2.5 over time (to show decrease over time)
- Amelia NY State hospitalizations-asthma and cvd (HAVE TO REPLACE HEART ATTACK WITH CVD)
)


There was limited data available for annual PM2.5 by NYC counies, so NY state-wide plots were selected to demonstrate the relationship between PM2.5 and hospitalizations.  



### Create PM2.5 Dataset, 2014
```{r message=FALSE, output=FALSE}
PM = read_csv("PM2.5_2014.csv") %>%
  janitor::clean_names() %>%
  #filter(borough == "Bronx" | borough == "Brooklyn" | borough == "Queens" | borough == "Staten Island" | borough == "Manhattan") %>%
  #filter(geo_type_name == "Borough") %>%
  #select(-geo_type_name, -geography, -geography_id, -indicator_name) %>%
  rename(PM_mean = mean_mcg_per_cubic_meter, 
         PM_tenth_percentile = x10th_percentile_mcg_per_cubic_meter,
         PM_ninety_percentile = x90th_percentile_mcg_per_cubic_meter, 
         PM_year = year,
         county_name = borough) %>%
  mutate(county_name = recode(county_name, "Brooklyn" = "Kings", "Staten Island" = "Richmond", "Manhattan" = "New York"))
```

### Create Children's Asthma Hospitalization Rate Dataset, 2012-14
```{r message=FALSE, output=FALSE}
asthma_hosp_child = read_csv("asthma_hosp_rate_10000_age0-17.csv") %>%
  janitor::clean_names() %>%
  filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years) %>%
  rename(asthma_hosp_percent_rate = percent_rate, 
         asthma_hosp_years = data_years)
```

### Create Age-Adjusted Cardiovascular Hospitalization Rates Dataset, 2012-14
```{r message=FALSE, output=FALSE}
aa_CVD_hosp = read_csv("age_adj_CVD_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(health_topic == "Cardiovascular Disease Indicators") %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percent_rate, data_years) %>%
  rename(aa_CVD_percent_rate = percent_rate, 
         aa_CVD_years = data_years)
```

### Create Age-Adjusted Heart Attack Hospitalization Rates Dataset, 2014
```{r message=FALSE, output=FALSE}
aa_heart_attack_hosp = read_csv("age_adj_heart_attack_hosp_rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(aa_hattack_percent_rate = percentage_rate_ratio,
         aa_hattack_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

### Create Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

### Create PM2.5 Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
PM_county_NYS = read_csv("PM2.5_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

### Plot PM2.5 (ug/m3) over time, 2000-2014 
## THIS ONE!
```{r}
PM_plot_conc_allyears = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Ambient PM2.5 Concentrations in New York State, 2000-2014", x = "Year", y = "PM2.5 (ug/m3)")
PM_plot_conc_allyears
```
### Plot PM2.5 (person days) over time, 2000-2014
```{r}
PM_plot_days_allyears = PM_county_NYS %>%
  filter(measure == "Person Days") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Time at Risk for Ambient PM2.5 Exposure in New York State, 2000-2014", x = "Year", y = "PM2.5 (person days)")
PM_plot_days_allyears
```


### Create Ozone Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
ozone_county_NYS = read_csv("ozone_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

### Plot Ozone over time, 2000-2014
```{r}
ozone_plot_allyears = ozone_county_NYS %>%
  filter(measure == "Person Days") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Time at Risk of Ambient Ozone Exposure in New York State, 2000-2014", x = "Year", y = "Ozone (person days)")
ozone_plot_allyears
```
### Plot Ozone per county, 2014
```{r}
ozone_plot_14 = ozone_county_NYS %>%
  filter(measure == "Person Days") %>%
  filter(year == "2014") %>%
  group_by(county) %>%
  ggplot(aes(x = reorder(county, -output), y = output, group = 1)) + geom_histogram(stat = "identity") + labs(title = "Time at Risk of Ambient Ozone Exposure in New York State, 2014", x = "County", y = "Ozone (person days)") + theme(axis.text.x = element_text(angle = 90))
ozone_plot_14
```

### Plot PM2.5 per county, 2014
```{r}
PM_plot_14 = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  filter(year == "2014") %>%
  group_by(county) %>%
  ggplot(aes(x = reorder(county, -output), y = output, group = 1)) + geom_histogram(stat = "identity") + labs(title = "Ambient PM2.5 Concentrations in New York State, 2014", x = "County", y = "PM2.5 (ug/m3)") + theme(axis.text.x = element_text(angle = 90))
PM_plot_14
```

### Plot Heart Attack Hospital Visits per County, 2014
```{r}
heart_plot_14 = aa_heart_attack_hosp %>% ##NEED TO CHANGE THIS DATASET
  ggplot(aes(x = reorder(county, -aa_hattack_percent_rate), y = aa_hattack_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Heart Attack Hospital Visit Rate per 10,000 by County in New York State, 2014")
heart_plot_14 
```

### Plot Asthma ER Admissions per County, 2014
```{r}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), y = asthma_ER_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Asthma ER Admission Rate (per 10,000)", title = "Asthma Emergency Room (ER) Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```






### Join heart attack, asthma, and PM datasets
redo join by county. not all counties overlap in all datasets!!!
```{r}
to_join_asthma_14 = asthma_ER %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex")
to_join_heart_14 = aa_heart_attack_hosp %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex")
  
to_join_ozone_14 = ozone_county_NYS %>%
  filter(year == "2014") %>%
  filter(county == "Bronx" | county == "New York" | county == "Kings" | county == "Erie" | county == "Richmond" | county == "Chautauqua" | county == "Queens" | county == "Suffolk" | county == "Albany" | county == "Monroe" | county == "Orange" | county == "Onondaga" | county == "Steuben" | county == "Essex") %>%
  rename(ozone_reading = output, ozone_unit = measure)
to_join_PM_14 = PM_county_NYS %>%
  filter (year == "2014") %>%
  rename (PM_reading = output, PM_unit = measure)
PM_ozone_joined = inner_join(to_join_ozone_14, to_join_PM_14, by = "county")
heart_asthma_joined = inner_join(to_join_heart_14, to_join_asthma_14, by = "county")
joined = inner_join(heart_asthma_joined, PM_ozone_joined, by = "county") %>%
  select(county, aa_hattack_percent_rate, asthma_ER_percent_rate, ozone_reading, ozone_unit, PM_reading, PM_unit)
```

### Plot Heart Attack & PM2.5, 2014
```{r}
heart_PM_plot = joined %>%  
  filter(PM_unit == "Micrograms/cubic meter (LC)") %>%
  group_by(county) %>%
  ggplot(aes(x = PM_reading, y = aa_hattack_percent_rate, color = county)) + geom_point() + labs(x = "", y = "", title = "by County in New York State, 2014") + labs(x = "Ambient PM2.5 (ug/m3)", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Association between PM2.5 and Heart Attack Hospital Visits in New York State, 2014")
heart_PM_plot
```

## Plot Asthma & PM2.5, 2014
```{r}
asthma_PM_plot = joined %>% 
  filter(PM_unit == "Micrograms/cubic meter (LC)") %>%
  group_by(county) %>%
  ggplot(aes(x = PM_reading, y = asthma_ER_percent_rate, color = county)) + geom_point() + labs(x = "Ambient PM2.5 (ug/m3)", y = "Asthma ER Admission Rate (per 10,000)", title = "Association between PM2.5 and Asthma Emergency Room (ER) Admissions in New York State, 2014")
asthma_PM_plot
```

## Ploto Heart Attack & Ozone, 2014
```{r}
heart_ozone_plot = joined %>%  
  filter(ozone_unit == "Person Days") %>%
  group_by(county) %>%
  ggplot(aes(x = ozone_reading, y = aa_hattack_percent_rate, color = county)) + geom_point() + labs(x = "Ozone (person days)", y = "Heart Attack Hospital Visit Rate (per 10,000)", title = "Association between Ozone and Heart Attack Hospital Visits in New York State, 2014")
heart_ozone_plot
```

## Plot Asthma & Ozone, 2014
```{r}
asthma_ozone_plot = joined %>% 
  filter(ozone_unit == "Person Days") %>%
  group_by(county) %>%
  ggplot(aes(x = ozone_reading, y = asthma_ER_percent_rate, color = county)) + geom_point() + labs(x = "Ozone (person days)", y = "Asthma ER Admission Rate (per 10,000)", title = "Association between Ozone and Asthma Emergency Room (ER) Admissions in New York State, 2014")
asthma_ozone_plot
```





ASHLEY'S CODE: 
Age-Adjusted Cardiovascular Disease Hospitalization Rate per 10,000 by County in NYC, 2012-2014

* Data source: NYSDOH Health Data

```{r load_clean_cvd, message = FALSE}
cvd_data = read_csv(file = "./data_AK/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate)
```

Note: Richmond County = Staten Island, Kings County = Brooklyn. We can rename from county to boroughs.

### Visualization of data

Asthma Emergency Department Visits in NYC, 2014 

```{r asthma_ed_bar}
asthma_ed_data %>%
  ggplot(aes(x = county_name, y = percentage_rate_ratio)) +
  labs(
    title = "Asthma Emergency Department Visits",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity", aes(fill = county_name)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw()
```

Asthma emergency department visits are highest in Bronx (substantially higher than the next highest county which is New York), while Richmond County (Staten Island) reports the fewest asthma emergency department visits.  

Age-Adjusted CVD Hospitalization Rate in NYC, 2012 - 2014 
```{r cvd_bar_nyc}
cvd_data %>%
  filter(county_name %in% c("Bronx", "Kings", "New York", "Queens", 
                            "Richmond")) %>% 
  ggplot(aes(x = county_name, y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity", aes(fill = county_name)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw()
```


Age-Adjusted CVD Hospitalization Rate in NY State, 2012 - 2014 

```{r cvd_bar_nys}
cvd_data %>%
  ggplot(aes(x = reorder(county_name, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")






