---
title: "Final Project Report"
author: "Anjile An, Ashley Kang, Amelia Grant-Alfieri, Divya Bisht"
date: 'Due December 11, 2018'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "90%"
)

library(tidyverse)
library(modelr)
library(mgcv)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Motivation

Unlike many factors we can control in determining our health, outdoor air quality is almost impossible to alter given its ubiquitous nature. We are interested in how environmental determinants of health, especially the short term effects of air quality measures, affect acute health outcomes. This also helps us understand the burden of acute health outcomes especially given our changing climate, and the potential impact of worsening air quality in both urban areas and rural areas. 

Our goal is to illustrate potential health disparities across New York State, the trends in the relationship between air quality and acute health outcomes, and areas (geographically and scientifically) requiring future research.

## Related work
*Anything that inspired you, such as a paper, a web site, or something we discussed in class.*


## Initial questions 
*What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?*

## Data
*Source, scraping method, cleaning, etc.*

## Exploratory analysis
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Background & Objectives 

Air pollution is known to have detrimental effects on human health. This study seeks to investigate air pollution, measured here as fine particulate matter (PM2.5), across New York State counties. In order to assess how PM2.5 exposure might be related to human health, we looked at asthma and cardiovascular (CVD) hospitalizations across New York counties. High PM2.5 causes damage to the respiratory system, which puts individuals at risk for health outcomes like asthma and heart disease. 

Article: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5260007/


### Exploration

Initial PM2.5 data exploration was focused on New York City counties. However, limited data was available available, so NY state county data were selected to demonstrate the relationship between PM2.5 and hospitalizations of asthma and CVD. 


Over 2000-2014, levels of PM2.5 across all counties in New York State have steadily decreased. 

#### PM2.5 Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
PM_county_NYS = read_csv("PM2.5_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

#### Plot PM2.5 (ug/m3) over time, 2000-2014 
```{r}
PM_plot_conc_allyears = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Ambient PM2.5 Concentrations in New York State, 2000-2014", x = "Year", y = "PM2.5 (ug/m3)")
PM_plot_conc_allyears
```


Next, we wanted to assess how PM2.5 exposure varied across counties in NY State. 

#### PM2.5 EPA dataset
```{r}
nyc_pm25 = read_csv(file = "./data_AA/annual_aqi_by_county_2014.csv") %>% 
  janitor::clean_names() %>%
  filter(state == "New York")
```

#### Days of PM2.5 by NY County
```{r}
pm_hist = nyc_pm25 %>%
  ggplot(aes(x = county, y = days_pm2_5)) + 
  labs(
    title = "Days of PM2.5 by Borough",
    x = "County",
    y = "Days with PM2.5") +
  geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90))

pm_hist
```


Literature shows that exposure to PM2.5 is associated with asthma and cardiovascular disease. Therefore, we wanted to assess hospitalization of asthma and CVD across counties. 

#### Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

#### Plot Asthma ER Admissions per County, 2014
```{r}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), y = asthma_ER_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Asthma ER Admission Rate (per 10,000)", title = "Asthma Emergency Room (ER) Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```

#### Cardiovascular hospitalizations
```{r load_clean_cvd, message = FALSE}
cvd_data = read_csv(file = "./data_AK/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate)
```

#### Age-Adjusted CVD Hospitalization Rate in NY State, 2012 - 2014 
```{r cvd_bar_nys}
cvd_data %>%
  ggplot(aes(x = reorder(county_name, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")
```


## Additional analysis
*If you undertake formal statistical analyses, describe these in detail.*

## Discussion
*What were your findings? Are they what you expect? What insights into the data can you make?*


