---
title: "Final Project Report"
author: "Anjile An, Ashley Kang, Amelia Grant-Alfieri, Divya Bisht"
date: 'Due December 11, 2018'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "90%"
)

library(tidyverse)
library(modelr)
library(mgcv)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Motivation

Unlike many factors we can control in determining our health, ambient (outdoor) air quality is almost impossible to alter given its ubiquitous nature. Across the globe, it accounts for significant morbidity and, to a lesser degree, mortality. Indoor air pollution is also an important exposure pathway particularly for women and children who live in households that burn solid fuel for cooking and heating. For this project, we focused on ambient air quality because there is more available data and because indoor air pollution is less widespread in the United States and New York State than in other places around the world. 

When someone breathes, they are exposed not to a single compound in isolation but rather to a mixture of compounds. Two compounds that are known to confer toxicity are ozone and fine particulate matter (PM2.5). Ozone is a fat soluble chemical than can bypass absorption in the upper respiratory system and penetrate down into the alveoli. PM2.5 is a tiny particle that, due to it's size, can also travel deep into the alveoli. Both PM2.5 and ozone can have harmful local effects in the respiratory system and, because of their ability to cross from the lung into the bloodstream, can have harmful distal effects throughout the cardiovascular system. 

We were interested in how air pollution in New York State, including specific compounds like PM2.5 and ozone, may lead to the acute exacerbation of chronic conditions like cardiovascular diseases and asthma as well as acute cardiovascular symptoms. Our goal is to illustrate potential health disparities across New York State, the trends in the relationship between air quality and acute health outcomes, and areas (geographically and scientifically) requiring future research.

## Related work
*Anything that inspired you, such as a paper, a web site, or something we discussed in class.*

<<<<<<< HEAD
Previous classwork utilizing NOAA data inspired us to explore other kinds of environmental data. We were also influenced by air pollution researchers at Columbia University like Dr. Frederica Perera, Dr. Marianthi-Anna Kioumourtzoglou, and others at the Center for Children's Environmental Health. 
=======
Previous classwork utilizing NOAA data inspired us to explore other kinds of environmental data.  
We were also influenced by air pollution researchers at Columbia University like Dr. Frederica Perera, Dr. Marianthi-Anna Kioumourtzoglou, and others at the Center for Children's Environmental Health. 

## Initial questions 
*What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?*

## Data
*Source, scraping method, cleaning, etc - Anjile*

Acknowledge multiple ways of measuring things - days, person-days, ug/m3 etc. Justify! 
- Ozone was not used bc not consistently measured across counties. 


## Exploratory analysis
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Background & Objectives 

Air pollution is known to have detrimental effects on human health. This study seeks to investigate air pollution, measured here as fine particulate matter (PM2.5), across New York State counties. In order to assess how PM2.5 exposure might be related to human health, we looked at asthma and cardiovascular (CVD) hospitalizations across New York counties. High PM2.5 causes damage to the respiratory system, which puts individuals at risk for health outcomes like asthma and heart disease. 

Article: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5260007/


### Exploration

Initial PM2.5 data exploration was focused on New York City counties. However, limited data was available available, so NY state county data were selected to demonstrate the relationship between PM2.5 and hospitalizations of asthma and CVD. 


Over 2000-2014, levels of PM2.5 across all counties in New York State have steadily decreased. 

#### PM2.5 Dataset, 2000-2014
```{r message=FALSE, output=FALSE}
PM_county_NYS = read_csv("PM2.5_county_NYS.csv") %>%
  janitor::clean_names() %>%
  #filter(year == "2014") %>%
  select(county_name, year, output, measure) %>%
  separate(county_name, into = c("county", "delete", sep = " ")) %>% 
  select(-delete) %>%
  mutate(county = recode(county, `New` = "New York"))
  #how to remove NA column?
```

#### Plot PM2.5 (ug/m3) over time, 2000-2014 
```{r}
PM_plot_conc_allyears = PM_county_NYS %>%
  filter(measure == "Micrograms/cubic meter (LC)") %>%
  group_by(county, year) %>%
  ggplot(aes(x = year, y = output, color = county)) + geom_line() + labs(title = "Ambient PM2.5 Concentrations in New York State, 2000-2014", x = "Year", y = "PM2.5 (ug/m3)")
PM_plot_conc_allyears
```


Next, we wanted to assess how PM2.5 exposure varied across counties in NY State. 

#### PM2.5 EPA dataset
```{r}
nyc_pm25 = read_csv(file = "./data_AA/annual_aqi_by_county_2014.csv") %>% 
  janitor::clean_names() %>%
  filter(state == "New York")
```

#### Days of PM2.5 by NY County
```{r}
pm_hist = nyc_pm25 %>%
  ggplot(aes(x = county, y = days_pm2_5)) + 
  labs(
    title = "Days of PM2.5 by Borough",
    x = "County",
    y = "Days with PM2.5") +
  geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90))

pm_hist
```


Literature shows that exposure to PM2.5 is associated with asthma and cardiovascular disease. Therefore, we wanted to assess hospitalization of asthma and CVD across counties. 

#### Asthma ER Admissions Rate Dataset, 2014
```{r message=FALSE, output=FALSE}
asthma_ER = read_csv("Asthma_ER_Rate_10000.csv") %>%
  janitor::clean_names() %>%
  #filter(county_name == "Bronx" | county_name == "Kings" | county_name == "Queens" | county_name == "Richmond" | county_name == "New York") %>%
  select(county_name, percentage_rate_ratio, data_years) %>%
  rename(asthma_ER_percent_rate = percentage_rate_ratio, 
         asthma_ER_years = data_years) %>%
  filter(county_name != "Long Island" & county_name != "New York City" & county_name != "Mid-Hudson" & county_name != "Capital Region" & county_name != "Mohawk Valley" & county_name != "North Country" & county_name != "Tug Hill Seaway" & county_name != "Central NY" & county_name != "Southern Tier" & county_name != "Finger Lakes" & county_name != "Western NY" & county_name != "New York State (excluding NYC)" & county_name != "New York State") %>% #to remove non-county regions
  rename(county = county_name)
```

#### Plot Asthma ER Admissions per County, 2014
```{r}
asthma_plot_14 = asthma_ER %>%
    ggplot(aes(x = reorder(county, -asthma_ER_percent_rate), y = asthma_ER_percent_rate, group = 1)) + geom_histogram(stat = "identity") + theme(axis.text.x = element_text(angle = 90)) + labs(x = "County", y = "Asthma ER Admission Rate (per 10,000)", title = "Asthma Emergency Room (ER) Admission Rate (per 10,000) by County in New York State, 2014")
asthma_plot_14 
```

#### Cardiovascular hospitalizations
```{r load_clean_cvd, message = FALSE}
cvd_data = read_csv(file = "./data_AK/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate)
```

#### Age-Adjusted CVD Hospitalization Rate in NY State, 2012 - 2014 
```{r cvd_bar_nys}
cvd_data %>%
  ggplot(aes(x = reorder(county_name, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")
```

=======
*Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.*
- Ashley, Div
>>>>>>> 86c08e47f37f82e7b6e7eb038c1fd090dfe96998

## Additional analysis
*If you undertake formal statistical analyses, describe these in detail.*
- Model building
- Looked at associations, trends

## Discussion
*What were your findings? Are they what you expect? What insights into the data can you make?*


Asthma is chronic disease so asthma emergency room visit rates reflect asthma exacerbation incidents. In contrast, cardiovascular hospitalization rates do not differentiate between a one-time acute episode like a stroke or an exacerbation of an underlying cardiovascular condition like high blood pressure or arrhythmia. 

Limitation
- So many things in air, mixtures, hard to get data on/complicated to analyze. Include in future models.

It is critical to note that poor communities are more susceptible to the potential life-threatening, acute effects of air pollution because they are more likely to have chronic diseases like asthma and heart disease and are more likely to live in more highly contaminated neighborhoods. 

