---
title: "Model Building"
author: "Anjile An"
date: 'December 4, 2018'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "90%"
)

library(tidyverse)
library(modelr)
library(mgcv)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Load and clean data

To build our model, we are looking at New York State asthma hospitalizations data, CVD hospitalizations data, and PM2.5 (air quality) data. In our exploratory analysis, we found that the trends of hospitalizations and PM2.5 do not trend together, and a potential confounder to that may be the number of hospitals in each county. 

```{r asthma_data, message=FALSE, output=FALSE}
# Asthma emergency department visit rate per 10,000 by county
# Source: NYSDOH Health Data NY
# remove "regions"

asthma_ed = read_csv(file = "./data_AA/asthma_ED_rate_10000.csv") %>% 
  janitor::clean_names() %>%
  select(county_name, event_count_rate, percentage_rate_ratio, data_years) %>%
  filter(!county_name %in% c("Capital Region", "Central NY", "Finger Lakes", "Long Island", "Mid-Hudson", "Mohawk Valley", "New York City", "New York State", "New York State (excluding NYC)", "North Country", "Southern Tier", "Tug Hill Seaway", "Western NY")) %>% 
  rename(asthma_hosp_percent_rate_ratio = percentage_rate_ratio, 
         asthma_event_count = event_count_rate,
         asthma_hosp_years = data_years,
         county = county_name)
```

```{r cvd_data, message=FALSE, output=FALSE}
# Age-adjusted cardiovascular disease hospitalization rate per 10,000 by county
# Source: NYSDOH Health Data NY
# remove "regions"

cvd_hosp = read_csv(file = "./data_AA/ageadjusted_cvd_hospitalization_rate_10000.csv") %>%
  janitor::clean_names() %>%
  filter(health_topic == "Cardiovascular Disease Indicators") %>%
  filter(!county_name %in% c("Capital Region", "Central NY", "Finger Lakes", "Long Island", "Mid-Hudson", "Mohawk Valley", "New York City", "New York State", "New York State (excluding NYC)", "North Country", "Southern Tier", "Tug Hill Seaway", "Western NY")) %>% 
  select(county_name, event_count, percent_rate, data_years) %>%
  rename(county = county_name,
         cvd_event_count = event_count,
         cvd_percent_rate = percent_rate,
         cvd_data_years = data_years)
```

```{r pm25_data, message=FALSE, output=FALSE}
# PM 2.5 annual summary data by county
# Source: US EPA AQS, 2014

nys_pm25 = read_csv(file = "./data_AA/annual_aqi_by_county_2014.csv") %>% 
  janitor::clean_names() %>%
  filter(state == "New York") %>%
  select(county, good_days:median_aqi) %>%
  rename(airq_good_days = good_days,
         airq_moderate_days = moderate_days,
         airq_unhealthy_for_sensitive_groups_days = unhealthy_for_sensitive_groups_days,
         airq_unhealthy_days = unhealthy_days,
         airq_very_unhealthy_days = very_unhealthy_days,
         airq_hazardous_days = hazardous_days)

# change NAs to 0 for ease of regression analysis
nys_pm25[is.na(nys_pm25)] = 0
```

```{r hosp_data, message=FALSE, output=FALSE}
# NYS county population and number of hospitals
# Source: NYS HealthData, 2014

num_hosp = read_csv(file = "./data_AA/nys_county_hospitals_2014.csv") %>% 
  janitor::clean_names() %>%
  rename(county = geography) %>%
  mutate(county = str_replace(county, " County", "")) %>%
  select(county, population, number_of_hospitals)
```

Joining the data:

```{r join_data}
asthma_cvd_join = full_join(asthma_ed, cvd_hosp, by = "county")
pm25_numhosp_join = full_join(nys_pm25, num_hosp, by = "county")

nys_joined = left_join(asthma_cvd_join, pm25_numhosp_join, by = "county")
```


**Checks before regression analysis:**

```{r reg_checks, message=FALSE, output=FALSE}
# Check for missing data

sum(is.na(asthma_ed)) 
sum(is.na(cvd_hosp))
sum(is.na(nys_pm25))
sum(is.na(num_hosp))
```

There are 2 pieces of missing data in `asthma_ed` file under Hamilton county, which could be a data entry error or due to the fact that Hamilton county has a very small population (4703) and none were hospitalized for asthma.

Visual checks were done in previous exploratory analysis. 


## Regression model

We will test two models each for asthma hospitalizations and CVD hospitalizations - the crude model and the adjusted model (including number of hospitals as a covariate).

Crude model:

> asthma = pm25 
> cvd = pm25

Adjusted model:

> asthma = pm25 + num_hosp
> cvd = pm25 + num_hosp

Interaction is not included, as it makes interpretation difficult.

#### Asthma model 

```{r asthma_reg_model, results = "hide"}
# Crude asthma model
crude_asthma_model = lm(asthma_hosp_percent_rate_ratio ~ total_unhealthy_days, data = nys_joined)
summary(crude_asthma_model)

# Adjusted asthma model
adj_asthma_model = lm(asthma_hosp_percent_rate_ratio ~ total_unhealthy_days + number_of_hospitals, data = nys_joined)
summary(adj_asthma_model)
```

#### CVD model

```{r cvd_reg_model, results = "hide"}
# Crude cvd model
crude_cvd_model = lm(cvd_percent_rate ~ total_unhealthy_days, data = nys_joined)
summary(crude_cvd_model)

# Adjusted cvd model
adj_cvd_model = lm(cvd_percent_rate ~ total_unhealthy_days + number_of_hospitals, data = nys_joined)
summary(adj_cvd_model)
```


## Cross validation

#### Asthma model

```{r asthma_cv}
set.seed(1)
cv_asthma = crossv_mc(nys_joined, 100) 

# Fit candidate models
options(warn = -1) # suppress printing all the warnings
cv_asthma = cv_asthma %>%
  mutate(crude_asthma_mod = map(train, ~lm(cvd_percent_rate ~ total_unhealthy_days, data = .x)),
         adj_asthma_mod = map(train, ~lm(cvd_percent_rate ~ total_unhealthy_days + number_of_hospitals, data = .x))) %>%
  mutate(rmse_crude = map2_dbl(crude_asthma_mod, test, ~rmse(model = .x, data = .y)),
         rmse_adj = map2_dbl(adj_asthma_mod, test, ~rmse(model = .x, data = .y)))

# Plot distribution of RMSE
cv_asthma %>% 
  select(starts_with("rmse")) %>% 
  gather(key = model, value = rmse) %>% 
  mutate(model = str_replace(model, "rmse_", ""),
         model = fct_inorder(model)) %>% 
  ggplot(aes(x = model, y = rmse)) + 
  labs(
    title = "Violin plots of RMSE, Asthma",
    y = "RMSE",
    x = "Model"
  ) +
  geom_violin()
```
