---
title: "Ashley's Exploratory Data Analysis"
author: "Ashley Kang"
date: "11/20/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(purrr)
library(broom)
library(rvest)
library(ggridges)
library(viridis)
library(patchwork)
```

### Loading and cleaning data

Annual Average of Fine Particulate Matter (PM 2.5) in NYC, 2001 - 2016 

* Data source: NYC DOHMH Environmental & Health Data

```{r load_clean_pm, message = FALSE}
pollution_data = read_csv(file = "./data_AK/Trends_in_Fine_Particulate_Matter_Annual_Average.csv") %>% 
  janitor::clean_names() %>%
  select(x_value, y_value) %>% 
  rename(year = x_value, PM_2.5 = y_value)
```

Annual Air Quality Index in NYC, 2012 - 2014

* Data source: US EPA

```{r load_clean_aqi, message = FALSE}
aqi_2012 = read_csv(file = "./data_AK/annual_aqi_by_county_2012.csv") %>% 
  janitor::clean_names() %>% 
  filter(state %in% "New York", 
         county %in% c("Bronx", "Kings", "New York", "Queens", 
                       "Richmond")) %>% 
  select(-state, -year)
  
aqi_2013 = read_csv(file = "./data_AK/annual_aqi_by_county_2013.csv") %>% 
  janitor::clean_names() %>% 
  filter(state %in% "New York", 
         county %in% c("Bronx", "Kings", "New York", "Queens", 
                       "Richmond")) %>% 
  select(-state, -year)

aqi_2014 = read_csv(file = "./data_AK/annual_aqi_by_county_2014.csv") %>% 
  janitor::clean_names() %>% 
  filter(state %in% "New York", 
         county %in% c("Bronx", "Kings", "New York", "Queens", 
                       "Richmond")) %>% 
  select(-state, -year)
```

Asthma Emergency Department Visit Rate per 10,000 by County in NYC, 2014

* Data source: NYSDOH Health Data

```{r load_clean_asthma_ed, message = FALSE}
asthma_ed_data = read_csv(file = "./data_AK/PA__Asthma_Emergency_Department_Visit_Rate_Per_10_000_by_County__Latest_Year.csv") %>% 
  janitor::clean_names() %>%
  select(county_name, event_count_rate, average_number_of_denominator_rate, 
         percentage_rate_ratio)
```

Age-Adjusted Cardiovascular Disease Hospitalization Rate per 10,000 by County in NYC, 2012-2014

* Data source: NYSDOH Health Data

```{r load_clean_cvd, message = FALSE}
cvd_data = read_csv(file = "./data_AK/Community_Health__Age-adjusted_Cardiovascular_Disease_Hospitalization_Rate_per_10_000_by_County_Map__Latest_Data.csv") %>% 
  janitor::clean_names() %>%
  filter(health_topic %in% "Cardiovascular Disease Indicators") %>% 
  select(county_name, event_count, average_number_of_denominator, 
         percent_rate)
```

Note: Richmond County = Staten Island, Kings County = Brooklyn. We can rename from county to boroughs.

### Visualization of data

Asthma Emergency Department Visits in NYC, 2014 

```{r asthma_ed_bar}
asthma_ed_data %>%
  ggplot(aes(x = county_name, y = percentage_rate_ratio)) +
  labs(
    title = "Asthma Emergency Department Visits",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity", aes(fill = county_name)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw()
```

Asthma emergency department visits are highest in Bronx (substantially higher than the next highest county which is New York), while Richmond County (Staten Island) reports the fewest asthma emergency department visits.  

Age-Adjusted CVD Hospitalization Rate in NYC, 2012 - 2014 
```{r cvd_bar_nyc}
cvd_data %>%
  filter(county_name %in% c("Bronx", "Kings", "New York", "Queens", 
                            "Richmond")) %>% 
  ggplot(aes(x = county_name, y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity", aes(fill = county_name)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw()
```


Age-Adjusted CVD Hospitalization Rate in NY State, 2012 - 2014 

```{r cvd_bar_nys}
cvd_data %>%
  ggplot(aes(x = reorder(county_name, -percent_rate), y = percent_rate)) +
  labs(
    title = "CVD Hospitalization Rate in NY State, 2012 - 2014",
    x = "County",
    y = "Hospitalizations per 10,000") +
  geom_histogram(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")

```

CVD hospitalizations are highest in Bronx, as well. However, the discrepency between the rate of CVD hospitalization in Bronx and Kings County (Brooklyn, which has the second highest rate) is not as extreme as it was for asthma emergency department visits. New York County reports the lowest rates of CVD hospitalization.  


Average Annual PM 2.5 in NYC, 2001 - 2016

```{r pm_graph}
pollution_data %>% 
  ggplot(aes(x = year, y = PM_2.5)) +
  geom_point() +
  geom_line() + 
  labs(
    title = "Average Annual PM 2.5 in NYC",
    x = "Year",
    y = "Average PM 2.5") +
  theme_bw()
```

Average annual PM 2.5 recordings were highest in 2001 and has generally decreased as we reach 2016. 

Number of Days with PM 2.5 Reading in NYC, 2012 - 2014

```{r aqi_pm_graph}
aqi_pm_2012_graph = aqi_2012 %>% 
  ggplot(aes(x = county, y = days_pm2_5)) +
  labs(
    title = "2012",
    x = "County",
    y = "Number of days") +
  geom_histogram(stat = "identity", aes(fill = county)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none")
  
aqi_pm_2013_graph = aqi_2013 %>% 
  ggplot(aes(x = county, y = days_pm2_5)) +
  labs(
    title = "2013",
    x = "County",
    y = "Number of days") +
  geom_histogram(stat = "identity", aes(fill = county)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none")

aqi_pm_2014_graph = aqi_2014 %>% 
  ggplot(aes(x = county, y = days_pm2_5)) +
  labs(
    title = "2014",
    x = "County",
    y = "Number of days") +
  geom_histogram(stat = "identity", aes(fill = county)) + 
  viridis::scale_fill_viridis(name = "County", discrete = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none")

aqi_pm_2012_graph + aqi_pm_2013_graph + aqi_pm_2014_graph
```

Kings County had the most number of days that recorded PM 2.5 levels across all three years.  


### Comments:

* Is it a problem if we have age-adjusted measures for CVD but not for asthma? More older people (compared to younger people) will be hospitalized for CVD, but do asthma hospitalizations vary by age?

* What do we do about the year range? CVD data is from 2012-2014, asthma data is from 2014, annual average readings of PM 2.5 is from 2001-2016.

* Haven't checked yet, but are there datasets with more years collected? Like from 2006 - 2014? It would be cool to explore whether asthma/CVD hospitalizations also decreased as PM 2.5 levels decreased over time.

* Potential limitation: no demographic data associated with hospitalization visits. Could SES, race, insurance status, etc. affect whether people visit the hospital even in cases of emergency?
